---
title: "ADA Final Project"
author: "Vicent Ssentumbwe"
date: '2022-11-22'
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r}
library(haven)
library(tidyverse)
library(foreign)
library(dplyr) # for manipulating the data frame
library(janitor) 
library(knitr)
library(ggplot2) #for graphs 
library(forcats)
library(stringr)
library(DescTools)
library(dataMaid)
library(tibble)
pacman::p_load(haven, tidyverse, naniar, VIM, mice, lattice, table1)
```

```{r}
Demographics <- read_sav('C:/Users/s.vicent/Documents/FALL_2022/Advanced data Analysis/Final_project/Teen-Demographic-SPSS.sav')
PhysicActivity <- read_sav('C:/Users/s.vicent/Documents/FALL_2022/Advanced data Analysis/Final_project/teenpa_public_updated.sav')

## Data can be accessed from https://cancercontrol.cancer.gov/brp/hbrb/flashe-study/flashe-files
```

```{r}
# merge two datasets by participant's ID
Merged_PA <- merge(Demographics,PhysicActivity,by="PID")
```


#Convert -9 (not assertained) to NA
```{r}
Merged_PA <- Merged_PA %>%
    mutate(across(
        TFOLFAMRUL:XTPREDWEEKSED,
        ~na_if(., -9)
    ))
```


## Pre-processing of data
```{r}
# Renaming the variables of interest
Merged_PA <- Merged_PA %>%
  rename (Age = TAGE,
          Sex = TSEX, 
          Race = TETHRAC_RC, 
          BMI = XTBMI, 
          Friend_support1 = TPSUPPRTPA, 
          Friend_support2 = TPNORMPA, 
          Parental_support1 = TPPPAENOUGH, 
          Parental_support2 = TPPPATKPLCS, 
          Parental_support3 = TPPPADECIDE, 
          Parental_support4 = TPPPAMKEXRC, 
          Parental_support5 = TPPPABEACTV, 
          Parental_support6 = TPPPAMKRULE, 
          Screen_Time1= TPPEDBADDAY, # To be reverse coded
          Screen_Time2= TPPEDDECIDE, # To be reverse coded
          Screen_Time3= TPPEDTKPLCS, # To be reverse coded
          Screen_Time4= TPPEDPARDEC, # To be reverse coded
          Screen_Time5= TPPEDPARREG, 
          Screen_Time6= TPPEDTRYLIM, 
          Screen_Time7= TPPEDMKRULE, # To be reverse coded
          Fac_indoor= TPANINDOOR, 
          Fac_School= TPANSCHREC, 
          Fac_Trails= TPANTRAILS,
          Fac_Playingfields= TPANPLAYFLD,
          Fac_Publicpark= TPANPUBPRK, 
          Env_stores= TPANSTORE, 
          Env_Crimes= TPCRIME, 
          Env_barriertraffic= TPTRAFFIC, 
          MVPA_atSchool= XTPREDMINMVPAS, 
          MVPA_outofSchool= XTPREDMINMVPAOFS, 
          MVPA_Weekend= XTPREDMINMVPAW
          )
```

```{r}
# Selecting  variables of interest
Merged_PA_Var <- select(Merged_PA, c('Age', 'Sex', 'Race', 'BMI', 'Friend_support1','Friend_support2', 'Parental_support1', 'Parental_support2', 'Parental_support3', 'Parental_support4', 'Parental_support5', 'Parental_support6', 'Screen_Time1', 'Screen_Time2', 'Screen_Time3', 'Screen_Time4', 'Screen_Time5', 'Screen_Time6', 'Screen_Time7', 'Fac_indoor', 'Fac_School', 'Fac_Trails', 'Fac_Playingfields', 'Fac_Publicpark', 'Env_stores', 'Env_Crimes', 'Env_barriertraffic', 'MVPA_atSchool', 'MVPA_outofSchool', 'MVPA_Weekend'))

head(Merged_PA_Var)
```
# Do a complete case analysis,Drop NAs-drop cases for any participant with missing value on any variables
```{r}
Merged_PA_exact <- Merged_PA_Var %>%
  drop_na()
```

```{r}
# Adding up parental support items
# Assessing the reliability of items in the scale using chronbac alpha

#install.packages("ltm")
library(ltm)
# 
Scale1 <- Merged_PA_exact[, c('Parental_support1','Parental_support2','Parental_support3','Parental_support4','Parental_support5','Parental_support6')]
#   
cronbach.alpha(Scale1) # Chronbach alpha is 0.831

# Adding up all items to get a composite score
Merged_PA_exact$Total_parental_support= Merged_PA_exact$Parental_support1+ Merged_PA_exact$Parental_support2+ Merged_PA_exact$Parental_support3+ Merged_PA_exact$Parental_support4+ Merged_PA_exact$Parental_support5+ Merged_PA_exact$Parental_support6
range (Merged_PA_exact$Total_parental_support) # Theoretical range is 6-30 and the exact range is also 6-30
summary(Merged_PA_exact$Total_parental_support)
```
# Getting the total available PA resources in the neighbour environmet
```{r}
Merged_PA_exact$PA_resources <- Merged_PA_exact$Fac_indoor +Merged_PA_exact$Fac_School +Merged_PA_exact$Fac_Trails +Merged_PA_exact$Fac_Playingfields +Merged_PA_exact$Fac_Publicpark
range (Merged_PA_exact$PA_resources) # Theoretical range is 0-5 and the exact range is also 0-5
summary(Merged_PA_exact$PA_resources)
```


```{r}
# Change class of categorical variable names to factor
Merged_PA_exact$Sex <- as.factor(Merged_PA_exact$Sex)
Merged_PA_exact$Race <- as.factor(Merged_PA_exact$Race)
Merged_PA_exact$Age <- as.factor(Merged_PA_exact$Age)
Merged_PA_exact$Env_stores <- as.factor(Merged_PA_exact$Env_stores)
Merged_PA_exact$Env_Crimes <- as.factor(Merged_PA_exact$Env_Crimes)
Merged_PA_exact$Env_barriertraffic <- as.factor(Merged_PA_exact$Env_barriertraffic)
Merged_PA_exact$Friend_support1 <- as.factor(Merged_PA_exact$Friend_support1)
Merged_PA_exact$Friend_support2 <- as.factor(Merged_PA_exact$Friend_support2)
```

```{r}
Merged_PA_exact<- Merged_PA_exact %>%
  mutate(Sex=recode_factor(Sex,
                           "1"= "Male",
                           "2" ="Female"))%>%
  mutate(Age = recode_factor(Age,
                              '2' = "12",
                              '3' = "13",
                              '4' = "14",
                              '5' = "15",
                              '6' = "16",
                              '7' = "17")) %>%
  mutate(Race=recode_factor(Race,
                              "1" ="Hispanic",
                              "2" = "Non-Hispanic Black or African American Only",
                              "3" ="Non-Hispanic White only",
                              "4" = "Non-Hispanic Other"))%>%
  
  mutate(Friend_support1 = recode_factor(Friend_support1,
                              '1' = "Strongly disagree",
                              '2' = "Somewhat disagree",
                              '3' = "Neither disagree nor agree",
                              '4' = "Somewhat agree",
                              '5' = "Strongly agree"))%>%
  
  mutate(Friend_support2 = recode_factor(Friend_support2,
                              '1' = "Strongly disagree",
                              '2' = "Somewhat disagree",
                              '3' = "Neither disagree nor agree",
                              '4' = "Somewhat agree",
                              '5' = "Strongly agree"))%>%
  
mutate(Env_stores = recode_factor(Env_stores,
                              '1' = "Strongly disagree",
                              '2' = "Somewhat disagree",
                              '3' = "Somewhat agree",
                              '4' = "Strongly agree"))%>%
  
mutate(Env_Crimes = recode_factor(Env_Crimes,
                              '1' = "Strongly disagree",
                              '2' = "Somewhat disagree",
                              '3' = "Somewhat agree",
                              '4' = "Strongly agree"))%>%
  
mutate(Env_barriertraffic = recode_factor(Env_barriertraffic,
                              '1' = "Strongly disagree",
                              '2' = "Somewhat disagree",
                              '3' = "Somewhat agree",
                              '4' = "Strongly agree"))
```

# Categorizing further the peer support variables

```{r}
Merged_PA_exact$Friend_support1 <- as.numeric(Merged_PA_exact$Friend_support1)
Merged_PA_exact$Friend_support1_R[Merged_PA_exact$Friend_support1>=1 & Merged_PA_exact$Friend_support1<=2]<-0
Merged_PA_exact$Friend_support1_R[Merged_PA_exact$Friend_support1==3]<-1
Merged_PA_exact$Friend_support1_R[Merged_PA_exact$Friend_support1>=4 & Merged_PA_exact$Friend_support1<=5]<-2

Merged_PA_exact$Friend_support2 <- as.numeric(Merged_PA_exact$Friend_support2)
Merged_PA_exact$Friend_support2_R[Merged_PA_exact$Friend_support2>=1 & Merged_PA_exact$Friend_support2<=2]<-0
Merged_PA_exact$Friend_support2_R[Merged_PA_exact$Friend_support2==3]<-1
Merged_PA_exact$Friend_support2_R[Merged_PA_exact$Friend_support2>=4 & Merged_PA_exact$Friend_support2<=5]<-2

Merged_PA_exact<- Merged_PA_exact %>%
mutate(Friend_support1_R = recode_factor(Friend_support1_R,
                              '0' = "Disagree",
                              '1' = "Neutral",
                              '2' = "Agree"))%>%
  
  mutate(Friend_support2_R = recode_factor(Friend_support2_R,
                              '0' = "Disagree",
                              '1' = "Neutral",
                              '2' = "Agree"))

```


# Categorizing further the items about the neighbourhood environment

```{r}
Merged_PA_exact<- Merged_PA_exact %>%
mutate(Env_stores_R = case_when(Env_stores %in% c("Strongly disagree","Somewhat disagree") ~ 0,
                             Env_stores %in% c("Somewhat agree","Strongly agree")~1)) %>%
mutate(Env_stores_R = factor(Env_stores_R, levels = c(0, 1), labels = c("Disagree", "Agree")))%>%


mutate(Env_Crimes_R = case_when(Env_Crimes %in% c("Strongly disagree","Somewhat disagree") ~ 0,
                             Env_Crimes %in% c("Somewhat agree","Strongly agree")~1)) %>%
mutate(Env_Crimes_R = factor(Env_Crimes_R, levels = c(0, 1), labels = c("Disagree", "Agree")))%>%
  

mutate(Env_barriertraffic_R = case_when(Env_barriertraffic %in% c("Strongly disagree","Somewhat disagree") ~ 0,
                             Env_barriertraffic %in% c("Somewhat agree","Strongly agree")~1)) %>%
mutate(Env_barriertraffic_R = factor(Env_barriertraffic_R, levels = c(0, 1), labels = c("Disagree", "Agree")))
```

# Categorizing further the PA resources in the neighboorhood
```{r}
Merged_PA_exact <- Merged_PA_exact %>%
  mutate(PA_resources_cat = case_when(PA_resources == 0  ~ 0,
                                       PA_resources >= 1 ~ 1),
         PA_resources_cat = factor(PA_resources_cat, 0:1, c("No PA resources", "Presence of PA resources"))) #make it as a factor variable
```


# Categorizing MVPA into two suffient and insufficient MVPA based on the CDC guidelines for adolescent MVPA (https://www.cdc.gov/healthyschools/physicalactivity/guidelines.htm#:~:text=Children%20and%20adolescents%20ages%206%20through%2017%20years%20should%20do,to%2Dvigorous%20physical%20activity%20daily.)
```{r}
Merged_PA_exact <- Merged_PA_exact %>%
  mutate(MVPA_atSchool_cat = case_when(MVPA_atSchool < 60  ~ 0,
                                       MVPA_atSchool >= 60 ~ 1),
         MVPA_atSchool_cat = factor(MVPA_atSchool_cat, 0:1, c("Insufficient MVPA", "Sufficient MVPA"))) #make it as a factor variable

Merged_PA_exact <- Merged_PA_exact %>%
  mutate(MVPA_outofSchool_cat = case_when(MVPA_outofSchool < 60  ~ 0,
                                       MVPA_outofSchool >= 60 ~ 1),
         MVPA_outofSchool_cat = factor(MVPA_outofSchool_cat, 0:1, c("Insufficient MVPA", "Sufficient MVPA"))) #make it as a factor variable

Merged_PA_exact <- Merged_PA_exact %>%
  mutate(MVPA_Weekend_cat = case_when(MVPA_Weekend < 60  ~ 0,
                                       MVPA_Weekend >= 60 ~ 1),
         MVPA_Weekend_cat = factor(MVPA_Weekend_cat, 0:1, c("Insufficient MVPA", "Sufficient MVPA"))) #make it as a factor variable
```

# FIGURE 1: Flow chart upto the analytical sample

```{r}
pacman::p_load(survival, survminer, tidyverse, readr, lmtest, table1)
```


```{r}
library(DiagrammeR)

#Figure 1
grViz("digraph flowchart {

      node [fontname = Helvetica, shape = rectangle, fontsize=10] 
      # notes indicate how many boxes you will have in your diagram. Since I have two sets of exclusions, I will have three nodes.
      tab1 [label = '@@1'] # starting
      tab2 [label = '@@2'] # exclusion 1
      tab3 [label = '@@3'] # exclusion 2
      tab4 [label = '@@4'] # Analytical sample
      
      tab1 -> tab2 -> tab3 -> tab4
}
      [1]: 'Number of participants who completed the demographics survey n=1737'
      [2]: 'Number of participants who completed physical activity survey, n= 1661)'
      [3]: 'Merged participants who completed both Physical activity survey and demographic survey n=1661'
      [4]: 'Exclude 370 participants with any missing data- analytical sample n=1291'
      ")
```


#TABLE1

```{r}
Tab1 <- table1(~Age + Sex + Race   + Total_parental_support  + Friend_support1_R + Friend_support2_R+ PA_resources_cat+Env_stores_R+Env_Crimes_R+Env_barriertraffic_R |MVPA_outofSchool_cat, overall="Total", rowlabelhead="Variable", Merged_PA_exact)
knitr::kable(Tab1)
```

# MODELING

```{r}
pacman::p_load(dplyr, ggplot2, car, odds.n.ends,readr, tidyverse, broom)
```

# First, visualize the distribution of data: Look at the distribution of Parental support and MVPA outside school
```{r}
  ggplot(Merged_PA_exact, aes(x = MVPA_outofSchool_cat, y = Total_parental_support, fill = MVPA_outofSchool_cat)) +
      geom_boxplot() +
      labs(x = "MVPA level", y = "Parental support") +
      theme_bw()

```
# univariable (unadjusted) logistic model for parental support.
```{r}
baseline_model1<-glm(MVPA_outofSchool_cat ~ Total_parental_support, data=Merged_PA_exact, family="binomial")
summary(baseline_model1)
```
# Get odds ratios
```{r}
odds.n.ends(baseline_model1)
```
# For every unit increase in parental support the odds of sufficient MVPA increase by 1.08 times. Model sensitivity is 50% and specificity is 68%

# Test lineality assumptions 
```{r}
Merged_PA_exact <- Merged_PA_exact %>%
  mutate(Total_parental_support_log = Total_parental_support * log(Total_parental_support)) # create term to test linearity

model2 <- glm(MVPA_outofSchool_cat~ Total_parental_support + Total_parental_support_log, data=Merged_PA_exact, family="binomial")

summary(model2)

# Since the p-value for the log term (Total_parental_support_log) is >.05, we fail to reject the hypothesis that the linearity assumption is not met using the Box Tidwell test, therefore we can use parental support as a continous or linear variable 
```
# Adding neighboorhood environment variables
```{r}
# All predictors- unadjusted
baseline_model3<-glm(MVPA_outofSchool_cat ~ Total_parental_support+Friend_support1_R+Friend_support2_R+PA_resources_cat+Env_stores_R+Env_Crimes_R+Env_barriertraffic_R, data=Merged_PA_exact, family="binomial")
summary(baseline_model3)
```
```{r}
odds.n.ends(baseline_model3)
```


# Adjusting for age, sex, and race
```{r}
Ajusted_model1<-glm(MVPA_outofSchool_cat ~ Total_parental_support+Friend_support1_R+Friend_support2_R+PA_resources_cat+Env_stores_R+Env_Crimes_R+Env_barriertraffic_R+Age+Sex+Race, data=Merged_PA_exact, family="binomial")
summary(Ajusted_model1)
```

```{r}
odds.n.ends(Ajusted_model1)
```
```{r}
odds.n.ends(Ajusted_model1, predProbPlot=TRUE, rocPlot = TRUE)
```


## Look at assumptions of multicollinearity using the vif function
```{r}
vif(Ajusted_model1)

# The VIF values are > 5, therefore there is no multicollinearity
```
## Lets look for influential observations using Cook's distance. First we will plot
```{r}
# Plot Cooks's Distance

plot(Ajusted_model1, which = 4, id.n = 3, col="blue") # which gives Cook's distance calculations (see https://stat.ethz.ch/R-manual/R-patched/library/stats/html/plot.lm.html)
```

## Use Cook's D calculations for each obervation and some other stats
```{r}
Ajusted_model1.data <- augment(Ajusted_model1) %>%  
  mutate(index = 1:n()) 

head(Ajusted_model1.data)
```

## Identify and exclude observations with a Cook's Distance greater than a cutoff, use 3 x the mean Cook's D 
```{r}
cuttoff <- mean(Ajusted_model1.data$.cooksd)*3 # defines cutoff variable

Merged_PA_exact_out <- Ajusted_model1.data %>% # excludes observations
  filter(.cooksd<cuttoff)

nrow(Merged_PA_exact_out)/nrow(Merged_PA_exact) # check percent kept
# 88.6% of the observations are kept
# many observations are being excluded, therefore it is not a good idea to remove outliers, all observations are retained.
```

# Try excluding influential observations using a defined cuttoff
```{r}
cuttoff <- 0.01 # defines cutoff variable

Merged_PA_exact_out2 <- Ajusted_model1.data %>% # excludes observations
  filter(.cooksd<cuttoff)

nrow(Merged_PA_exact_out2)/nrow(Merged_PA_exact) # check percent kept
```
## Rerun model and compare coefficients
```{r}
Ajusted_model2<-glm(MVPA_outofSchool_cat ~ Total_parental_support+Friend_support1_R+Friend_support2_R+PA_resources_cat+Env_stores_R+Env_Crimes_R+Env_barriertraffic_R+Age+Sex+Race, data=Merged_PA_exact_out2, family="binomial")
summary(Ajusted_model2)
odds.n.ends(Ajusted_model2)

```
# compare coefficients
```{r}
compareCoefs(Ajusted_model1, Ajusted_model2)

# After excluding the outliers, the model performance doesnot change a lot therefore all observations are retained.
```
# Assess model fitness of Adjusted_model1
```{r}
blr_model_fit_stats(Ajusted_model1)
blr_test_hosmer_lemeshow(Ajusted_model1)

# The Hosmer lemeshow goodness of fit test indicates a good model fitness (p-value<0.05)
```
#Testing effect modification of sex on the relationship between social support on MVPA

```{r}
Effect_mod_model<-glm(MVPA_outofSchool_cat ~ Total_parental_support+Friend_support1_R+Friend_support2_R+PA_resources_cat+Env_stores_R+Env_Crimes_R+Env_barriertraffic_R+Age+Sex+Race+Total_parental_support*Sex+Friend_support2_R*Sex+Friend_support2_R*Sex, data=Merged_PA_exact, family="binomial")

summary(Effect_mod_model)
odds.n.ends(Effect_mod_model)

Ajusted_model3<-glm(MVPA_outofSchool_cat ~ Total_parental_support+Friend_support1_R+Friend_support2_R+PA_resources_cat+Env_stores_R+Env_Crimes_R+Env_barriertraffic_R+Age+Sex+Race, data=Merged_PA_exact, family="binomial")
summary(Ajusted_model3)
odds.n.ends(Ajusted_model3)
```

```{r}
lrtest(Effect_mod_model, Ajusted_model3)
lrtest(Ajusted_model3,Effect_mod_model)

# No effect modification by sex (P>0.05)
```





